---
title: "Dimension Reduction"
author: "Yuting Sun"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r}
#Load required libraries
library(readr)
library(tidyverse)
library(stopwords)
library(wordcloud)
library(FactoMineR)
library(factoextra)
```
## Task 1 Build the shape table

```{r}
#load the dataset
ufo <- read.csv("data/nuforc_sightings.csv")
```

```{r}
#Check country variations starting with 'U'
ufo %>%
  filter(startsWith(country, "U")) %>%
  count(country, sort = TRUE)
```
Multiple entries for "USA" (e.g., "Usa", "usa") indicate data cleaning is needed.


```{r}
#Clean and standardize country data
ufo %>%
  mutate(country_upper = toupper(trimws(country))) %>%
  filter(grepl("US|AMERICA|STATES|UNITED STATES|U.S.", country_upper)) %>%
  count(country, sort = TRUE)


```
Should excluded ambiguous entries like "Bahamas/USA" to focus on mainland USA.
also excluded U.S. Virgin Islands.


```{r}
#Verify no missing country values
sum(is.na(ufo$country))

#check missing values of state in the USA
usa_no_state <- ufo %>%
  filter((country == "USA" | country == "US") & (is.na(state) | state == "" | state == "Unkown"))

nrow(usa_no_state)
```

```{r}
# Filter for USA and check state completeness
ufo_us <- ufo %>%
  mutate(
    country = toupper(trimws(country))
  ) %>%
  
  filter(country == "USA")


```

```{r}
# Filter for USA and check state completeness
ufo_us %>%
  count(country, sort = TRUE)

```


```{r}
# Clean and summarize state data
ufo_us %>%
  count(state, sort = TRUE) %>%
  head(100)

```
some enteies' state is not USA's state, and some did not use abbreviation, etc
```{r}
#Define valid U.S. states
us_states <- c("AL", "AK", "AZ", "AR", "CA", "CO", "CT", "DE", "FL", "GA",
               "HI", "ID", "IL", "IN", "IA", "KS", "KY", "LA", "ME", "MD",
               "MA", "MI", "MN", "MS", "MO", "MT", "NE", "NV", "NH", "NJ",
               "NM", "NY", "NC", "ND", "OH", "OK", "OR", "PA", "RI", "SC",
               "SD", "TN", "TX", "UT", "VT", "VA", "WA", "WV", "WI", "WY",
               "DC")
#Create state mapping (abbreviation to full name)
state_mapping <- data.frame(
  abbreviation = c("AL", "AK", "AZ", "AR", "CA", "CO", "CT", "DE", "FL", "GA",
                   "HI", "ID", "IL", "IN", "IA", "KS", "KY", "LA", "ME", "MD",
                   "MA", "MI", "MN", "MS", "MO", "MT", "NE", "NV", "NH", "NJ",
                   "NM", "NY", "NC", "ND", "OH", "OK", "OR", "PA", "RI", "SC",
                   "SD", "TN", "TX", "UT", "VT", "VA", "WA", "WV", "WI", "WY", "DC"),
  full_name = c("Alabama", "Alaska", "Arizona", "Arkansas", "California", 
                "Colorado", "Connecticut", "Delaware", "Florida", "Georgia",
                "Hawaii", "Idaho", "Illinois", "Indiana", "Iowa", "Kansas",
                "Kentucky", "Louisiana", "Maine", "Maryland", "Massachusetts",
                "Michigan", "Minnesota", "Mississippi", "Missouri", "Montana",
                "Nebraska", "Nevada", "New Hampshire", "New Jersey", "New Mexico",
                "New York", "North Carolina", "North Dakota", "Ohio", "Oklahoma",
                "Oregon", "Pennsylvania", "Rhode Island", "South Carolina",
                "South Dakota", "Tennessee", "Texas", "Utah", "Vermont",
                "Virginia", "Washington", "West Virginia", "Wisconsin", "Wyoming",
                "District of Columbia")
)
```




```{r}
# unique state in ufo_us
all_states_in_data <- unique(ufo_us$state)

# match with state_mapping
non_us_states <- all_states_in_data[!all_states_in_data %in% us_states]

print("not US states：")
print(non_us_states)


ufo_us %>%
  filter(state %in% non_us_states) %>%
  count(state, sort = TRUE)
```
```{r}
# Clean state data
ufo_us_clean <- ufo_us %>%
  mutate(
    # tolower
    state = trimws(tolower(state)),
    
    # transfer
    state = case_when(
      state == "montana" ~ "MT",
      state == "new york" ~ "NY",
      state == "ohio" ~ "OH",
      state == "west virginia" ~ "WV",
      state == "wisconsin" ~ "WI",
      TRUE ~ toupper(state)  
    )
  ) %>%
  # filter states not in the US
  filter(state %in% us_states)

# 
print(paste("state count:", length(unique(ufo_us_clean$state))))

# checking states
unique(ufo_us_clean$state) %>% sort()
```



```{r}
# data quality of shape
unique(ufo_us_clean$shape)

ufo_us_clean %>% 
  count(shape, sort = TRUE)
```

```{r}
# Is there NA, Null, trimming spaces/white space in shape column
sum(is.na(ufo_us_clean$shape))
sum(ufo_us_clean$shape == "", na.rm = TRUE)
head(ufo_us_clean$shape, 20)
```
```{r}
#Clean shape column
ufo_us_clean_shape <- ufo_us_clean %>%
  mutate(
    
    shape = tolower(shape),
   
    shape = trimws(shape),
    
   
    shape = case_when(
      is.na(shape) ~ "unknown",
      shape == "" ~ "unknown",
      TRUE ~ shape 
    )
  )


ufo_us_clean_shape %>%
  count(shape, sort = TRUE)
```


```{r}
# Create shape-by-state frequency table
shape_by_state <- ufo_us_clean_shape %>%
  count(state, shape) %>%           
  pivot_wider(
    names_from = shape,             # shape-column
    values_from = n,                      # count
    values_fill = 0                       
  )

# Check dimensions and unique states
head(shape_by_state)
dim(shape_by_state)  
unique(shape_by_state$state)

# Identify state with most "Circle" sightings
shape_by_state$state[which.max(shape_by_state$circle)]

```
There are 25 distinct shapes categories.
CA reported the highest number of "Circle" sightings(1637).



## Task 2 PCA on the shape table

```{r}
#convert counts to proporities
shape_matrix <- shape_by_state %>%
  column_to_rownames("state")
state_totals <- rowSums(shape_matrix)
shape_proportions <- shape_matrix / state_totals # Row-normalization
head(rowSums(shape_proportions))
```

```{r}
#Perform PCA
pca_result <- prcomp(shape_proportions, scale. = TRUE, center = TRUE)

#Summary of PCA results
summary(pca_result)
```

```{r}
#Create screen plot
fviz_eig(pca_result, 
         addlabels = TRUE, 
         ylim = c(0, 50),
         main = "Scree Plot: Variance Explained by Principal Components",
         xlab = "Principal Components",
         ylab = "Percentage of Explained Variance")
```
<br>
The plot shows that the first few principal components explain a substantial portion of the variance.


```{r}
#Create scree plot
fviz_pca_ind(pca_result,
             col.ind = "contrib", 
             gradient.cols = c("#00AFBB", "#E7B800", "#FC4E07"),
             repel = TRUE, 
             labelsize = 3,
             pointsize = 2,
             title = "PCA - States Colored by Contribution to PC1-PC2",
             xlab = "PC1",
             ylab = "PC2")
```
<br>
Based on the scatterplot, there are no visible regional clusters.  
However, the plot clearly shows two significant outliers, DC and ND, which have unique UFO shape distributions compared to all other states.

```{r}
# Loadings for PC1 and PC2
loadings <- pca_result$rotation[, 1:2]
loadings_df <- as.data.frame(loadings)
loadings_df$shape <- rownames(loadings_df)
```


```{r}
# Bar plots for loadings
ggplot(loadings_df, aes(x = reorder(shape, PC1), y = PC1)) +
  geom_bar(stat = "identity", fill = "steelblue") +
  coord_flip() +
  labs(title = "PC1 Loadings: Shape Contributions",
       x = "UFO Shape",
       y = "Loading on PC1") +
  theme_minimal()
```

```{r}
ggplot(loadings_df, aes(x = reorder(shape, PC2), y = PC2)) +
  geom_bar(stat = "identity", fill = "coral") +
  coord_flip() +
  labs(title = "PC2 Loadings: Shape Contributions",
       x = "UFO Shape",
       y = "Loading on PC2") +
  theme_minimal()
```


```{r}
#Biplot
fviz_pca_biplot(pca_result,
                repel = TRUE,
                col.var = "red", 
                col.ind = "blue", 
                title = "PCA Biplot: States and UFO Shapes",
                xlab = "PC1",
                ylab = "PC2")
```

```{r}
#Contribution plots for PC1
fviz_contrib(pca_result, choice = "var", axes = 1, top = 10,
             title = "Top 10 Shapes Contributing to PC1")
```
<br>
Based on the fviz_contrib() bar plot, the shapes that contribute the most to PC1 are:Triangle, Light and Diamond.

```{r}
#Contribution plots for PC2
fviz_contrib(pca_result, choice = "var", axes = 2, top = 10,
             title = "Top 10 Shapes Contributing to PC2")
```
<br>
Based on the fviz_contrib() bar plot, the shapes that contribute the most to PC1 are:Unkown, Orb and Circle.

## Task 3 Clean and tokenize the summaries


```{r}
#Clean summary
ufo_us_clean_shape <- ufo_us_clean_shape %>%
  mutate(
    summary_clean = ifelse(is.na(summary) | summary == "", NA, summary),  # Handle NA or empty
    summary_clean = tolower(summary),  # Convert to lowercase
    summary_clean = gsub("[^[ -~]]", "", summary_clean),  # Remove non-ASCII characters
    summary_clean = trimws(summary_clean),  # Trim whitespace from edges
    summary_clean = gsub("\\s+", " ", summary_clean)  # Replace repeated internal whitespace with single space
  )

print(head(ufo_us_clean_shape$summary_clean, 10))  # Debug output
  
```





```{r}
    #install.packages("tidytext")
```



```{r}
# Tokenize the cleaned summaries into words using tidytext
library(tidytext)
tokens <- ufo_us_clean_shape %>%
  unnest_tokens(word, summary_clean)  # Break into array of words

# Generate a table showing the most frequent words (top 20 for brevity)
word_freq <- tokens %>%
  count(word, sort = TRUE)

head(word_freq, 20)
```

```{r}
# Create a word cloud
wordcloud(words = word_freq$word[1:100], freq = word_freq$n[1:100], 
          min.freq = 100, random.order = FALSE, colors = brewer.pal(8, "Dark2"))
```
<br>
In this initial output, the most frequent words are mostly common, such as "the", "a", "in", "and", "of" etc. These do not provide much insight into the UFO sightings themselves, as they are stopwords that appear in almost any English text.


```{r}
# Remove stopwords using the stopwords package
stop_words <- stopwords("en")  # Get English stopwords

tokens_no_stop <- tokens %>%
  filter(!word %in% stop_words)  # Remove stopwords

# Re-generate the frequency table (top 20)
word_freq_no_stop <- tokens_no_stop %>%
  count(word, sort = TRUE)

head(word_freq_no_stop, 20)
```

```{r}
# Re-make the word cloud without stopwords
wordcloud(words = word_freq_no_stop$word[1:100], freq = word_freq_no_stop$n[1:100], 
          min.freq = 100, random.order = FALSE, colors = brewer.pal(8, "Dark2"))


```
<br>
After stopword removal, the words that feel most characteristic of these reports include terms like  "lights", "light", "sky", "object","bright", "moving", etc. These suggest common themes in descriptions, such as visual appearances (colors, brightness) and behaviors (movement, hovering) of the sighted objects.

## Task 4 Build the keyword table and repeat PCA

```{r}
# Define a vocabulary of the top 100 most frequent words from cleaned summaries
vocab <- word_freq_no_stop %>%
  filter(nchar(word) >= 3) %>%  # At least 3 characters
  slice(1:100) %>%  # Top 100
  pull(word)
```

```{r}
# Create a wide table: state-by-keyword counts
# First, filter tokens to only include words in vocab
tokens_vocab <- tokens_no_stop %>%
  filter(word %in% vocab)

# count per state and pivot wider
keyword_by_state <- tokens_vocab %>%
  group_by(state, word) %>%
  summarise(n = n(), .groups = "drop") %>%
  pivot_wider(
    names_from = word,
    values_from = n,
    values_fill = 0
  )
```


```{r}
#show dimensions and head
dim(keyword_by_state)
head(keyword_by_state)
```

```{r}
#PCA
#convert counts to proporities
keyword_matrix <- keyword_by_state %>%
  column_to_rownames("state")
state_totals_keywords <- rowSums(keyword_matrix)
keyword_proportions <- keyword_matrix / state_totals_keywords
head(rowSums(keyword_proportions))
```

```{r}
# Perform PCA
pca_result_keywords <- prcomp(keyword_proportions, scale. = TRUE, center = TRUE)
summary(pca_result_keywords)
```

```{r}
# Scree plot
fviz_eig(pca_result_keywords, 
         addlabels = TRUE, 
         ylim = c(0, 50),
         main = "Scree Plot: Variance Explained by Principal Components (Keywords)",
         xlab = "Principal Components",
         ylab = "Percentage of Explained Variance")
```
<br>
The scree plot shows that the first few principal components explain a substantial portion of the variance, similar to the shape PCA.

```{r}
# Scatterplot of states on PC1-PC2
fviz_pca_ind(pca_result_keywords,
             col.ind = "contrib", 
             gradient.cols = c("#00AFBB", "#E7B800", "#FC4E07"),
             repel = TRUE, 
             labelsize = 3,
             pointsize = 2,
             title = "PCA - States Colored by Contribution to PC1-PC2 (Keywords)",
             xlab = "PC1",
             ylab = "PC2")
```
<br>
Similar to the shape PCA, there may be no strong regional clusters, but outliers like DC, HI appear again if their keyword usage are unique. Other states could cluster based on similar descriptive language in reports.

```{r}
# Examine loadings for PC1 and PC2
loadings_keywords <- pca_result_keywords$rotation[, 1:2]
loadings_df_keywords <- as.data.frame(loadings_keywords)
loadings_df_keywords$word <- rownames(loadings_df_keywords)
```
```{r}
# Bar plot for PC1 loadings
ggplot(loadings_df_keywords, aes(x = reorder(word, PC1), y = PC1)) +
  geom_bar(stat = "identity", fill = "steelblue") +
  coord_flip() +
  labs(title = "PC1 Loadings: Keyword Contributions",
       x = "Keyword",
       y = "Loading on PC1") +
  theme_minimal()
```
```{r}
# Bar plot for PC2 loadings
ggplot(loadings_df_keywords, aes(x = reorder(word, PC2), y = PC2)) +
  geom_bar(stat = "identity", fill = "coral") +
  coord_flip() +
  labs(title = "PC2 Loadings: Keyword Contributions",
       x = "Keyword",
       y = "Loading on PC2") +
  theme_minimal()
```

```{r}
# Biplot
fviz_pca_biplot(pca_result_keywords,
                repel = TRUE,
                col.var = "red", 
                col.ind = "blue", 
                title = "PCA Biplot: States and Keywords",
                xlab = "PC1",
                ylab = "PC2")
```


```{r}
# Top 10 contributing to PC1
fviz_contrib(pca_result_keywords, choice = "var", axes = 1, top = 10,
             title = "Top 10 Keywords Contributing to PC1")
```
<br>
Based on the fviz_contrib() bar plot, keywords contribute the most to PC1 are:obversed, sighting, line.

```{r}
# Top 10 contributing to PC2
fviz_contrib(pca_result_keywords, choice = "var", axes = 2, top = 10,
             title = "Top 10 Keywords Contributing to PC2")
```
<br>
Based on the fviz_contrib() bar plot, keywords contribute the most to PC2 are: bright, craft, shaped

Compare with shape-based PCA:
Similarities: Both analyses identify DC and HI as outliers, suggesting that their UFO sighting reports may feature distinctive characteristics. This could be attributed to DC not being a traditional state but a federal district and HI being far away from USA mainland, which might influence the nature and reporting of observations. Even with normalization across states, these unique aspects persist, indicating inherent differences in thieir data.
Differences: Keywords provide more nuanced patterns (e.g., colors, movements) beyond just shapes, potentially revealing thematic differences in how sightings are described across states, whereas shapes are more categorical.
